---
title: "Coursera Regression Models Course_Project1"
author: "doyougnu"
date: "September 19, 2015"
output: pdf_document
---
##Executive Summary
In this 
##Data loading and first look
First lets load and look at the data to assess variation by variable, I will
also do a quick plot to see a linear regression of each variable with mpg

```{r dataLoad, echo = TRUE, fig.height= 5, fig.width= 8, message=FALSE}
data(mtcars)
mtcars <- mtcars #call mtcars because data load is lazy

#load libs
library(ggplot2); library(dplyr); library(tidyr)

#some quick munging
df <- mtcars %>% gather(carvar, var.value, cyl:carb)
ggplot(df)  + 
    geom_point(aes(x = var.value, y = mpg, color = carvar)) +
    facet_wrap( ~ carvar, scales = "free_x")
```

Looks like most variables are well correlated to mpg. For a good model we need
factors that explain the variation in MPG while being orthogonal to other 
explanatory factors. I will refer to domain specific knowledge provided in the
Henderson and Vellermans 1981 paper referenced in the mtcars dataset[1]. This is
an important tactic because in that paper the authors provide a physical 
explanation of work of a car (work in the physics sense of force applied over a
distance). I will be relying on their results heavily for this model.

##Model Adjustments
In that paper the authors state:
1. GPM (Gallon per Mile) is a more useful response variable due to the 
non-linearity of several variables, Weight, Displacement etc.. and domain
specific knowledge of the sources of work in a car 
2. Given GPM, Weight is the single greatest predictor in a linear regression
model(Henderson et al. 397)
3. Calculating the ratio of Horse Power to Weight provides a useful factor to 
add to the meodel because it provides a metric to asses how overpowered a car is
in this dataset (defined by large HP/WT). Furthermore, it is uncorrelated to 
WEIGHT (r = 0.03) (Henderson et al. 398)

##Model Construction
In this analysis we are interested in the difference between standard and 
automatic transmissions, given the considerations above we will want to look at
the difference between automatic and manual transmissions **holding all other
significant variables constant**. Thus we need to add the trans variable to the
model described by Henderson et al

```{r modelCons, echo = TRUE}
df <- mtcars; df$GPM <- mtcars$mpg ^ -1; df$HPWT <- mtcars$hp / mtcars$wt
fit <- lm(GPM ~ wt + HPWT + factor(am), data = df)
summary(fit)$coeff
```

As one can see from above the addition of a transmission variable is 
insignificant - which is unsurprising because most of the variation in the data
is already explained by WEIGHT and HP/WT. However this is only true **for this
dataset* looking at the cars listed in the dataset we are unlikely to find cars
whose mpg would be significantly altered by manual and automatic transmitions. 
For a more detailed analysis on why see Henderson et al, page 396-397. Thus,
for this dataset, neither automatic nor manual transmission is better for MPG 
and furthermore the difference between the two is negligible.

##Conclusion
In this short analysis I set out to determine two things 1. Does transmission 
type matter in reference to MPG and 2. Is so, what is the quantitative 
difference between the two in terms of MPG. I found that 1. Given prior work on 
the subject, the variation in this dataset is explained by Weight of the vehicle
and the Horse Power to Weight ratio. 2. The addition of transmission to the model
was statistically insignificant which as noted in prior research on this dataset
is most likely due to the absence of mid sized sedans in the dataset.

###Links
1: http://www.mortality.org/INdb/2008/02/12/8/document.pdf